\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-05-30 paper.qmd
Update section lifting model to p-D

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{}   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in 2D represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@cfa69] and multidimensional scaling (MDS) [@borg2005]. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)
<!--### Notation -->

@@

With our goal being to make a representation of this 2D layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbour information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first hexagonally bin the data, reducing it from $n$ to $m\leq n$ observations, and connect the bin centroids. This process serves to reduce some noisiness in the resulting surface shown in \pD{}. The steps in this process are shown in @fig-NLDR-scurve, and documented below.

@@

<!--We deliberately separate out the creation of the hexagonal grid from the mapping of points on the grid.-->
#### Summarization of the data in 2D

#### Generating edges to indicate neighbors and removing long edges

@@

The last step is to lift the \kD{} model into \pD{} by computing \pD{} vectors that represent bin centroids. Define the set $H^k$ to be the set of points that belong to bin $k$. We use the \pD{} Euclidean mean of the points in $H^k$ to map the centroid $C_{h}^{(k)} = (c_{h1}, c_{h2})$ to a point in \pD{}. Let the $i$-th component of the \pD{} mean be

$$f_i = \frac{1}{\left|H^k\right|}\sum_{y \in H^k} y_i,$$

with corresponding \pD{} vector $\vec{f_i}$. Then, $f(u(x))$ maps a \pD{} point to the \pD{} model estimate, completing the model cycle.

Furthermore, edges that exist between \kD{} representations should also generate edges in \pD{} by connecting \pD{} mapping of the corresponding \kD{} representations.

@@

To assess how well our method captures and represents the underlying structure of the \pD{} data, residuals are important. Residuals are computed by taking the squared \pD{} Euclidean distance, $\sum_{i = 1}^{n}\sum_{j = 1}^{p} (x_{ij} - C^{(p)}_{x_ij})^2$. Additionally, Mean Squared Error (MSE), $\sum_{i = 1}^{n}\sum_{j = 1}^{p} \frac{(x_{ij} - C^{(p)}_{x_ij})^2}{n}$, is computed based on residuals.

@@

```{r}
#| echo: false
#| label: fig-bins-scurve
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the UMAP applied to S-curve data: (a) $b = 12$ (3, 5), (b) $b = 190$ (10, 19), and (c) $b = 2176$ (34, 64). The hexbins are colored based on the density of points, with darker colors indicating higher density and yellow color representing lower density within each bin. What is the number of bins that would be effective?"
#| fig-width: 6
#| fig-height: 3

@@

benchmark_label_df <- tibble(y = benchmark_scurve - 0.03,
                             x = 1.4,
                             label = paste0("benchmark is ", benchmark_scurve))

>>>>

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{}   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@cfa69] and multidimensional scaling (MDS) [@borg2005]. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)
<!--### Notation -->

@@

With our goal being to make a representation of this \gD{}    layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbour information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first hexagonally bin the data, reducing it from $n$ to $m\leq n$ observations, and connect the bin centroids. This process serves to reduce some noisiness in the resulting surface shown in \pD{}. The steps in this process are shown in @fig-NLDR-scurve, and documented below.

@@

<!--We deliberately separate out the creation of the hexagonal grid from the mapping of points on the grid.-->

#### Indicating neighbors by line segments connecting centroids

@@

The last step is to lift the \kD{} model into \pD{} by computing \pD{} vectors that represent bin centroids. We use the \pD{} mean of the points in $H_h$ to map the centroid $C_{h}^{(k)} = (c_{h1}, c_{h2}, \dots, c_{hk})$ to a point in \pD{}. Let the \pD{} mean be

$$C_{h}^{(p)} = \frac{1}{n_h}\sum_{i =1}^{n_h} x_i, h = {1, \dots, b; n_h > 0}.$$
Furthermore, line segments that exist in the \kD{} model generate line segments in \pD{} by connecting the \pD{} means of the corresponding \kD{} bin centroids.

@@

To assess how well our method captures and represents the underlying structure of the \pD{} data, residuals are important. Residuals are computed by taking the squared \pD{} Euclidean distance, $\sum_{i = 1}^{n}\sum_{j = 1}^{p} (x_{ij} - C^{(p)}_{hx_ij})^2$. Additionally, Mean Squared Error (MSE), $\frac{1}{n}\sum_{i = 1}^{n}\sum_{j = 1}^{p}(x_{ij} - C^{(p)}_{hx_ij})^2$, is computed based on residuals.

@@

```{r}
#| echo: false
#| label: fig-bins-scurve
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the UMAP applied to S-curve data: (a) $b = 15$ (3, 5), (b) $b = 190$ (10, 19), and (c) $b = 2176$ (34, 64). The hexbins are colored based on the density of points, with darker colors indicating higher density and yellow color representing lower density within each bin. What is the number of bins that would be effective?"
#| fig-width: 6
#| fig-height: 3

@@

benchmark_label_df <- tibble(y = benchmark_scurve - 0.03,
                             x = 1.3,
                             label = paste0("benchmark is ", benchmark_scurve))

### 2024-05-30 paper.tex
Update section lifting model to p-D

At first glance, thinking of NLDR as a modeling technique might seem
strange. It is a simplified representation or abstraction of a system,
process, or phenomenon in the real world. The \pD{} observations are the
realization of the phenomenon, and the \kD{} NLDR layout is the
simplified representation. From a statistical perspective we can
consider the distances between points in the \kD{} layout to be variance
that the model explains, and the (relative) difference with their
distances in \pD{} is the error, or unexplained variance. We can also
imagine that the positioning of points in 2D represent the fitted
values, that will have some prescribed position in \pD{} that can be
compared with their observed values. This is the conceptual framework
underlying the more formal versions of factor analysis \citep{cfa69} and
multidimensional scaling (MDS) \citep{borg2005}. (Note that, for this
thinking the full \pD{} data needs to be available, not just the
interpoint distances.)

@@

\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, k, m$ & number of observations, variables, embedding dimension, number of non-empty bins, respectively\\
$\mathbfit{X}, \mathbfit{x}$ & $p$-dimensional data (population, sample)\\
$\mathbfit{y}$ & $k$-dimensional layout\\
$P$ & orthonormal basis, generating a $d\text{-}dimensional$ linear projection of $p$-dimensional data\\
$T$ & true  model\\
\addlinespace
$g$ & functional mapping from \pD{} to \kD{}, especially as prescribed by NLDR\\
$\mathbfit{\theta}$ & (Hyper-) parameters for NLDR method\\
$r$ & ranges of the embedding components\\
$C^{(j)}$ & $j$-dimensional bin centers\\
$(b_1, b_2)$ & number of bins in each direction\\
\addlinespace
$(a_1, a_2)$ & binwidths, distance between centroids in each direction\\
$(s_1, \ s_2)$ & starting coordinates of the hexagonal grid\\
$q$ & buffer to ensure hexgrid covers data, proportion of data range, 0-1\\
$b$, $b'$ & total and non-empty hexagon bins in the grid\\
$n_m$ & number of observations within the $m^{th}$ hexagon\\
\addlinespace
$h$ & hexagonal id\\
\bottomrule
\end{tabular}
\endgroup{}

@@

With our goal being to make a representation of this 2D layout that can
be lifted into high-dimensional space, the layout needs to be augmented
to include neighbour information. A simple approach would be to
triangulate the points and add edges. A more stable approach is to first
hexagonally bin the data, reducing it from \(n\) to \(m\leq n\)
observations, and connect the bin centroids. This process serves to
reduce some noisiness in the resulting surface shown in \pD{}. The steps
in this process are shown in Figure~\ref{fig-NLDR-scurve}, and
documented below.

@@

\subsubsection{Summarization of the data in
2D}\label{summarization-of-the-data-in-2d}

@@

\subsubsection{Generating edges to indicate neighbors and removing long
edges}\label{generating-edges-to-indicate-neighbors-and-removing-long-edges}

@@

The last step is to lift the \kD{} model into \pD{} by computing \pD{}
vectors that represent bin centroids. Define the set \(H^k\) to be the
set of points that belong to bin \(k\). We use the \pD{} Euclidean mean
of the points in \(H^k\) to map the centroid
\(C_{h}^{(k)} = (c_{h1}, c_{h2})\) to a point in \pD{}. Let the \(i\)-th
component of the \pD{} mean be

\[f_i = \frac{1}{\left|H^k\right|}\sum_{y \in H^k} y_i,\]

with corresponding \pD{} vector \(\vec{f_i}\). Then, \(f(u(x))\) maps a
\pD{} point to the \pD{} model estimate, completing the model cycle.

Furthermore, edges that exist between \kD{} representations should also
generate edges in \pD{} by connecting \pD{} mapping of the corresponding
\kD{} representations.

@@

To assess how well our method captures and represents the underlying
structure of the \pD{} data, residuals are important. Residuals are
computed by taking the squared \pD{} Euclidean distance,
\(\sum_{i = 1}^{n}\sum_{j = 1}^{p} (x_{ij} - C^{(p)}_{x_ij})^2\).
Additionally, Mean Squared Error (MSE),
\(\sum_{i = 1}^{n}\sum_{j = 1}^{p} \frac{(x_{ij} - C^{(p)}_{x_ij})^2}{n}\),
is computed based on residuals.

@@

\caption{\label{fig-bins-scurve}Hexbin plots from different number of
bins for the UMAP applied to S-curve data: (a) \(b = 12\) (3, 5), (b)
\(b = 190\) (10, 19), and (c) \(b = 2176\) (34, 64). The hexbins are
colored based on the density of points, with darker colors indicating
higher density and yellow color representing lower density within each
bin. What is the number of bins that would be effective?}

>>>>

At first glance, thinking of NLDR as a modeling technique might seem
strange. It is a simplified representation or abstraction of a system,
process, or phenomenon in the real world. The \pD{} observations are the
realization of the phenomenon, and the \kD{} NLDR layout is the
simplified representation. From a statistical perspective we can
consider the distances between points in the \kD{} layout to be variance
that the model explains, and the (relative) difference with their
distances in \pD{} is the error, or unexplained variance. We can also
imagine that the positioning of points in \gD{} represent the fitted
values, that will have some prescribed position in \pD{} that can be
compared with their observed values. This is the conceptual framework
underlying the more formal versions of factor analysis \citep{cfa69} and
multidimensional scaling (MDS) \citep{borg2005}. (Note that, for this
thinking the full \pD{} data needs to be available, not just the
interpoint distances.)

@@

\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, k$ & number of observations, variables, embedding dimension, respectively\\
$\mathbfit{X}, \mathbfit{x}$ & $p$-dimensional data (population, sample)\\
$\mathbfit{y}$ & $k$-dimensional layout\\
$P$ & orthonormal basis, generating a $d\text{-}dimensional$ linear projection of $p$-dimensional data\\
$T$ & true  model\\
\addlinespace
$g$ & functional mapping from \pD{} to \kD{}, especially as prescribed by NLDR\\
$\mathbfit{\theta}$ & (Hyper-) parameters for NLDR method\\
$r$ & ranges of the embedding components\\
$C^{(j)}$ & $j$-dimensional bin centers\\
$(b_1, b_2)$ & number of bins in each direction\\
\addlinespace
$(a_1, a_2)$ & binwidths, distance between centroids in each direction\\
$(s_1, \ s_2)$ & starting coordinates of the hexagonal grid\\
$q$ & buffer to ensure hexgrid covers data, proportion of data range, 0-1\\
$m$ & number of non-empty bins\\
$b$ & number of  hexagons in the grid\\
\addlinespace
$h$ & hexagonal id\\
\bottomrule
\end{tabular}
\endgroup{}

@@

With our goal being to make a representation of this \gD{} layout that
can be lifted into high-dimensional space, the layout needs to be
augmented to include neighbour information. A simple approach would be
to triangulate the points and add edges. A more stable approach is to
first hexagonally bin the data, reducing it from \(n\) to \(m\leq n\)
observations, and connect the bin centroids. This process serves to
reduce some noisiness in the resulting surface shown in \pD{}. The steps
in this process are shown in Figure~\ref{fig-NLDR-scurve}, and
documented below.

@@

@@

\subsubsection{Indicating neighbors by line segments connecting
centroids}\label{indicating-neighbors-by-line-segments-connecting-centroids}

@@

The last step is to lift the \kD{} model into \pD{} by computing \pD{}
vectors that represent bin centroids. We use the \pD{} mean of the
points in \(H_h\) to map the centroid
\(C_{h}^{(k)} = (c_{h1}, c_{h2}, \dots, c_{hk})\) to a point in \pD{}.
Let the \pD{} mean be

\[C_{h}^{(p)} = \frac{1}{n_h}\sum_{i =1}^{n_h} x_i, h = {1, \dots, b; n_h > 0}.\]
Furthermore, line segments that exist in the \kD{} model generate line
segments in \pD{} by connecting the \pD{} means of the corresponding
\kD{} bin centroids.

@@

To assess how well our method captures and represents the underlying
structure of the \pD{} data, residuals are important. Residuals are
computed by taking the squared \pD{} Euclidean distance,
\(\sum_{i = 1}^{n}\sum_{j = 1}^{p} (x_{ij} - C^{(p)}_{hx_ij})^2\).
Additionally, Mean Squared Error (MSE),
\(\frac{1}{n}\sum_{i = 1}^{n}\sum_{j = 1}^{p}(x_{ij} - C^{(p)}_{hx_ij})^2\),
is computed based on residuals.

@@

\caption{\label{fig-bins-scurve}Hexbin plots from different number of
bins for the UMAP applied to S-curve data: (a) \(b = 15\) (3, 5), (b)
\(b = 190\) (10, 19), and (c) \(b = 2176\) (34, 64). The hexbins are
colored based on the density of points, with darker colors indicating
higher density and yellow color representing lower density within each
bin. What is the number of bins that would be effective?}
